Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2014-11-20 manuscript.tex
Reminder to include a section on unit tests
Reminder to add subsect: gzz x height and time x D
Describe the division figure in the text

\documentclass[paper,twocolumn]{geophysics}
%\documentclass[manuscript]{geophysics}
%\documentclass[manuscript,endfloat]{geophysics}
\usepackage{graphicx}
\usepackage{amsmath}
\newcommand{\mitbf}[1]{\mathbf{{#1}}}

@@

\begin{equation}
    d = \left[
        r^2 + r_2^2 - 2 r r_2
        \cos(\sin\phi\sin\Delta\phi + \cos\phi\cos\Delta\phi
            \cos(\lambda - \Delta\lambda))
        \right],
    \label{eq:distance}
\end{equation}

@@

@@

Because of this stacks are also known as ``Last-in-first-out'', or LIFO,
structures.

@@

We must define a maximum size for the stack.

@@

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last element from it).
        This will cause \texttt{stktop} to be reduced by one.
    \item Compute the distance $d$ between the tesseroid and
        the computation point (equation~\ref{eq:distance}).
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ (equations~\ref{eq:sizelon}-\ref{eq:sizer}).
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.
    \item If none of the dimensions fail the above condition,
        compute the gravitational effect of the tesseroid using the
        Gauss-Legendre Quadrature with order two
        (equation~\ref{eq:glq3d}).
    \item If any of the dimensions fail the condition:
    \begin{enumerate}
        \item Divide the tesseroid in half along each dimension that failed
             the condition.
        \item Check if the number new (smaller) tesseroids fit in the stack.
             In other words, the number of new elements plus \texttt{stktop}
             must be smaller than the maximum stack size.
             If the stack is full, warn the user of a ``stack overflow''
             and compute the effect of the larger tesseroid (step 5).
        \item If the stack is not full, place the smaller tesseroids into
             the stack.
    \end{enumerate}
    \item Repeat the above steps until the stack is empty
        (\texttt{stktop = -1}).
\end{enumerate}

@@

\begin{figure}
    \centering
    \includegraphics{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $D$ equal to
        (b) 1, (c) 2, and (d) 6.
        $d$ is the Cartesian distance between
        the center of the top face of the tesseroid
        and $P$.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $D$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
    }
    \label{fig:division}
\end{figure}

@@

>>>>

%\documentclass[paper,twocolumn]{geophysics}
\documentclass[manuscript]{geophysics}
%\documentclass[manuscript,endfloat]{geophysics}
\usepackage{graphicx}
\usepackage{amsmath}
\newcommand{\mitbf}[1]{\mathbf{{#1}}}

@@

\begin{equation}
    d = \left[
        r^2 + r_2^2 - 2 r r_2
        \cos(\sin\phi\sin\Delta\phi + \cos\phi\cos\Delta\phi
            \cos(\lambda - \Delta\lambda))
        \right]^{\frac{1}{2}},
    \label{eq:distance}
\end{equation}

@@

USE THE MAX OF SOUTH AND NORTH FOR DLON? MIGHT MAKE IT BETTER IN THE POLES.

@@

Stacks are also known as ``Last-in-first-out'', or LIFO, structures.

@@

We must define a maximum size for the stack to allocate memory for it.

Defining a maximum size allows us to avoid an infinite loop and large
consumption of computer memory in case the computation point is
on (or sufficiently close to) the surface of the tesseroid.

@@

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last element from it).
        This will cause \texttt{stktop} to be reduced by one.
    \item Compute the distance $d$ between the tesseroid and
        the computation point (equation~\ref{eq:distance}).
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ (equations~\ref{eq:sizelon}-\ref{eq:sizer}).
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.
    \item If none of the dimensions fail the above condition,
        compute the gravitational effect of the tesseroid using the
        Gauss-Legendre Quadrature with order two
        (equation~\ref{eq:glq3d}).
    \item If any of the dimensions fail the condition:
    \begin{enumerate}
        \item Divide the tesseroid in half along each dimension that failed
             the condition.
        \item Check if the number of new elements plus \texttt{stktop}
             is smaller than the maximum stack size.
             If it is not (i.e., the stack is full),
             warn the user of a ``stack overflow''
             and compute the effect of the larger tesseroid (step 5).
        \item If the stack is not full, place the smaller tesseroids into
             the stack.
    \end{enumerate}
    \item Repeat the above steps until the stack is empty
        (\texttt{stktop} is equal to -1).
\end{enumerate}

@@

Figure~\ref{fig:division} shows the effect of the distance-size ration $D$
in the adaptive discretization of a tesseroid for the computation at a point P.

Computing the gravitational effect of the tesseroid in
Figure~\ref{fig:division}a using the algorithm described above
is equivalent to computing the effect of the tesseroids in
Figures~\ref{fig:division}b-d.

Note that increasing the value of $D$ results in a finer mesh of tesseroids.

For $D = 1$ (Figure~\ref{fig:division}b) the number of tesseroids in the
adaptive discretization is 12,
for $D = 2$ (Figure~\ref{fig:division}c) it is 52,
and for $D = 6$ (Figure~\ref{fig:division}d) it is 1092.

\begin{figure}
    \centering
    \includegraphics{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $D$ equal to
        (b) 1, (c) 2, and (d) 6.
        $d$ is the Cartesian distance between
        the center of the top face of the tesseroid
        and $P$.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $D$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
        The number of tesseroids in the discretization is (b) 12, (c) 52,
        and (d) 1092.
    }
    \label{fig:division}
\end{figure}

@@

\subsection{$g_{zz}$ variability with computation height}

\subsection{Computational performance as a function of $D$}

For close P, vary $D$ and graph computation time.
